Sparse spectral and $p$-finite element methods on disk-slices and trapeziums I: zero boundary conditions

### 2019-04-02 papers/halfdisk/halfdisk.tex
References to eqns and figs

We can then obtain the matrix operator for the Laplacian \(\Delta\), that will take us from coefficients for expansion in the weighted space
\[
\bigWii(x,y) = \Wii(x,y) \: \bigPii(x,y)
\]
to coefficients in the non-weighted space
\[
\bigPii(x,y).
\]
Importantly, this operator will have Banded-Block-Banded structure, and hence will be sparse, as seen in FIGURE. Note that this construction will ensure the imposition of the Dirichlet zero boundary conditions on $\Omega$. The matrix operator for the Laplacian we denote $\laplacewii$ acting on the coefficients vector is then given by
\begin{align}
    \laplacewii := D_x^{(0,0)} \: W_x^{(1,1)} + T^{(0,1)\to(1,1)} \: D_y^{(0,0)} \: T_W^{(1,0)\to(0,0)} \: W_y^{(1,1)}.
\end{align}

Further, we can also obtain the matrix operator for the Laplacian $\Delta$ that will take us from coefficients for expansion in the space
$$
\bigPoo(x,y)
$$
to coefficients in the space
$$
\bigP^{(2,2)}(x,y).
$$
Importantly, again, this operator will have Banded-Block-Banded structure, and hence will be sparse, as seen in FIGURE. We denote this matrix operator that acts on the coefficients vector as $\laplaceoo$, and is given by
\begin{align}
    \laplaceoo := D_x^{(1,1)} \: D_x^{(0,0)} + T^{(1,2)\to(2,2)} \: D_y^{(1,1)} \: T^{(0,1)\to(1,1)} \: D_y^{(0,0)}.
\end{align}

We can use the same technique for constructing a sparse Laplacian operator to construct the Biharmonic operator $\Delta^2$, that ensures we use a basis that satisfies both zero Dirichlet and Neumann conditions. On this note, we can further represent the Laplacian as a map from coefficients in the space 
$$
\bigW^{(2,2)}
$$
to coefficients in the space
$$
\bigPoo.
$$
Note that a function expanded in the $\bigW^{(2,2)}$ basis will satisfy both zero Dirichlet and Neumann boundary conditions on $\Omega$. We denote this matrix operator as $\laplacewtt$, and is given by
\begin{align}
	\laplacewtt := W_x^{(1,1)} \: W_x^{(2,2)} + T_W^{(1,0)\to(0,0)} \: W_y^{(1,1)} \: T_W^{(2,1)\to(1,1)} \: W_y^{(2,2)}.
\end{align}
The sparsity and structure of this matrix can be seen in FIGURE.

The sparsity of the these operators can be shown using a simple integration by parts argument, and can be seen in FIGURE.

@@

\begin{figure}
	\centering
	\includegraphics[scale=0.4]{solutionblocknorms}
	\caption{The norms of each block of the computed solution of the Poisson equation with the given right hand side functions.}
	\centering
	\label{fig:solutionblocknorms}
\end{figure}

Find \(u(x,y)\) given a function \(f(x,y)\) such that:
\begin{align}
	\begin{cases}
    		\Delta u(x,y) = f(x,y) \quad \text{in } \Omega \\
		u(x,y) = 0 \quad \text{on } \partial \Omega
	\end{cases}.
\end{align}
noting the imposition of zero Dirichlet boundary conditions on $u$.

We can tackle the problem as follows. Fix \(a, b = 1,1\). Denote the coefficient vector for expansion of $u$ in the $\bigWii$ OP basis up to degree $N$ by $\mathbf{u}$, and the coefficient vector for expansion of $f$ in the $\bigPii$ OP basis up to degree $N$ by $\mathbf{f}$. Since $f$ is known, we can obtain $\mathbf{f}$ using the quadrature rule above. In matrix-vector notation, our system hence becomes:
\begin{align}
    \laplacewii \mathbf{u} = \mathbf{f}
\end{align}
which can be solved to find $\mathbf{u}$.
In FIGURE we see the solution to the Poisson equation with zero boundary conditions given in EQN in the half-disk $\Omega$. In FIGURE we show the norms of each block of calculated coefficients of the approximation for four right-hand sides of the Poisson equation with N = 999????. The rate of decay in the coefficients is a proxy for the rate of convergence of the computed solution. \bstodo{Describe what plot tells us}

%\begin{figure}
%	\includegraphics[scale=0.4]{example4coeffs}
%\centering
%\caption{Plot of the difference between the previous solution's coefficients and the current solution's coefficients under increasing degree of approximation (number of coefficients used) for $N = 2,\dots,10$, using RHS of $f(x,y) = 2(1-x^2-y^2)\cos(x) - 8x\sin(x) - x(1-x^2-y^2)\sin(x) - 4x^2 \cos(x) - 12xy$ for the Poisson equation. The solution is converging as seen by the decreasing (absolute value) difference of coefficient values.}
%\centering
%\label{fig:poisson}
%\end{figure}

\begin{figure}
	\centering
	\includegraphics[scale=0.4]{sparsityoflaplacian}
        \label{fig:sparsity}
    	\caption{A "spy" plot of the $\laplacewii$ matrix operator for the Laplacian showing its sparsity.}
	\centering
\end{figure}

@@

Find \(u(x,y)\) given functions $v$, $f : \Omega \to \R$ such that:
\begin{align}
	\begin{cases}
    		\Delta u(x,y) + k^2 v(x,y) \; u(x,y) = f(x,y) \quad \text{in } \Omega \\
		u(x,y) = 0 \quad \text{on } \partial \Omega
	\end{cases}.
\end{align}
where $k \in \R$, noting the imposition of zero Dirichlet boundary conditions on $u$.

We can tackle the problem as follows. Fix \(a, b = 1,1\). Denote the coefficient vector for expansion of $u$ in the $\bigWii$ OP basis up to degree $N$ by $\mathbf{u}$, and the coefficient vector for expansion of $f$ in the $\bigPii$ OP basis up to degree $N$ by $\mathbf{f}$. Since $f$ is known, we can obtain $\mathbf{f}$ using the quadrature rule above. We can obtain the matrix operator for the variable-coefficient function $v(x,y)$ by using the Clenshaw algorithm with matrix inputs as the Jacobi matrices ${J_x^{(0,0)}}^T, {J_y^{(0,0)}}^T$, yielding an operator matrix of the same dimension as the input Jacobi matries. We can denote the resulting operator acting on coefficients in the $\bigPoo$ space by $V({J_x^{(0,0)}}^T, {J_y^{(0,0)}}^T)$. In matrix-vector notation, our system hence becomes:
\begin{align}
    (\laplacewii + k^2 T^{(0,0)\to(1,1)} \: V({J_x^{(0,0)}}^T, {J_y^{(0,0)}}^T) \: T_W^{(1,1)\to(0,0)}) \mathbf{u} = \mathbf{f}
\end{align}
which can be solved to find $\mathbf{u}$. We can see the sparsity and structure of this matrix system in FIGURE with $v(x,y) = xy^2$ as an example. In FIGURE we see the solution to the inhomogenous variable-coefficient Helmholtz equation with zero boundary conditions given in EQN in the half-disk $\Omega$, with  $v(x,y) = xy^2$ and $f(x,y) = xy e^x$.

@@

\begin{figure}
	\centering
	\includegraphics[scale=0.3]{solution}
        	\label{fig:solution}
    	\caption{The computed solution to $\Delta u = f$ with zero boundary conditions with $f(x,y) = 1 + \text{erf}(5(1 - 10((x - 0.5)^2 + y^2)))$.}
	\centering
\end{figure}

@@

Find \(u(x,y)\) given a function \(f(x,y)\) such that:
\begin{align}
	\begin{cases}
    		\Delta^2 u(x,y) = 0 \quad \text{in } \Omega \\
		u(x,y) = 0, \quad \frac{\partial u}{\partial n}(x,y) = 0 \quad \text{on } \partial \Omega
	\end{cases}.
\end{align}
where $\Delta^2$ is the Biharmonic operator, noting the imposition of zero Dirichlet and Neumann boundary conditions on $u$. We can construct a matrix operator for $\Delta^2$ that will take coefficients in the space
$$
\bigW^{(2,2)}
$$
to coefficients in the space
$$
\bigP^{(2,2)}.
$$
Note that any function expanded in the $\bigW^{(2,2)}$ basis will satisfy both zero Dirichlet and zero Neumann boundary conditions on $\Omega$. The matrix operator for the Biharmonic operator is then given by
\begin{align}
	\biharmonic = \laplaceoo \: \laplacewtt.
\end{align}
In FIGURE we see the solution to the Biharmonic equation EQN in the half-disk $\Omega$.

>>>>

\begin{figure}
	\centering
	\includegraphics[scale=0.4]{sparsityoflaplacian}
    	\caption{A "spy" plot of the $\laplacewii$ matrix operator for the Laplacian showing its sparsity.}
        \label{fig:sparsity}
	\centering
\end{figure}

We can then obtain the matrix operator for the Laplacian \(\Delta\), that will take us from coefficients for expansion in the weighted space
\[
\bigWii(x,y) = \Wii(x,y) \: \bigPii(x,y)
\]
to coefficients in the non-weighted space
\[
\bigPii(x,y).
\]
Importantly, this operator will have Banded-Block-Banded structure, and hence will be sparse, as seen in Figure \ref{fig:sparsity}. Note that this construction will ensure the imposition of the Dirichlet zero boundary conditions on $\Omega$. The matrix operator for the Laplacian we denote $\laplacewii$ acting on the coefficients vector is then given by
\begin{align}
    \laplacewii := D_x^{(0,0)} \: W_x^{(1,1)} + T^{(0,1)\to(1,1)} \: D_y^{(0,0)} \: T_W^{(1,0)\to(0,0)} \: W_y^{(1,1)}.
\end{align}

Further, we can also obtain the matrix operator for the Laplacian $\Delta$ that will take us from coefficients for expansion in the space
$$
\bigPoo(x,y)
$$
to coefficients in the space
$$
\bigP^{(2,2)}(x,y).
$$
Importantly, again, this operator will have Banded-Block-Banded structure, and hence will be sparse, as seen in Figure \ref{fig:sparsity}. We denote this matrix operator that acts on the coefficients vector as $\laplaceoo$, and is given by
\begin{align}
    \laplaceoo := D_x^{(1,1)} \: D_x^{(0,0)} + T^{(1,2)\to(2,2)} \: D_y^{(1,1)} \: T^{(0,1)\to(1,1)} \: D_y^{(0,0)}.
\end{align}

We can use the same technique for constructing a sparse Laplacian operator to construct the Biharmonic operator $\Delta^2$, that ensures we use a basis that satisfies both zero Dirichlet and Neumann conditions. On this note, we can further represent the Laplacian as a map from coefficients in the space 
$$
\bigW^{(2,2)}
$$
to coefficients in the space
$$
\bigPoo.
$$
Note that a function expanded in the $\bigW^{(2,2)}$ basis will satisfy both zero Dirichlet and Neumann boundary conditions on $\Omega$. We denote this matrix operator as $\laplacewtt$, and is given by
\begin{align}
	\laplacewtt := W_x^{(1,1)} \: W_x^{(2,2)} + T_W^{(1,0)\to(0,0)} \: W_y^{(1,1)} \: T_W^{(2,1)\to(1,1)} \: W_y^{(2,2)}.
\end{align}
The sparsity and structure of this matrix can be seen in Figure \ref{fig:sparsity}.

The sparsity of the these operators can be shown using a simple integration by parts argument.

@@

\begin{figure}
	\begin{subfigure}[t]{0.3\textwidth}
	\includegraphics[scale=0.3]{solution}
	\centering
	\caption{}
        	\label{fig:solution-poisson}
	\end{subfigure}
	\begin{subfigure}[t]{0.4\textwidth}
        	\includegraphics[scale=0.4]{solutionblocknorms}
	\centering
	\caption{}
        	\label{fig:solutionblocknorms}
	\end{subfigure}
	\caption{(a) The computed solution to $\Delta u = f$ with zero boundary conditions with $f(x,y) = 1 + \text{erf}(5(1 - 10((x - 0.5)^2 + y^2)))$. (b) The norms of each block of the computed solution of the Poisson equation with the given right hand side functions.}
	\centering
\end{figure}

Find \(u(x,y)\) given a function \(f(x,y)\) such that:
\begin{align}
	\begin{cases}
    		\Delta u(x,y) = f(x,y) \quad \text{in } \Omega \\
		u(x,y) = 0 \quad \text{on } \partial \Omega
	\end{cases}.
	\label{eqn:poisson}
\end{align}
noting the imposition of zero Dirichlet boundary conditions on $u$.

We can tackle the problem as follows. Fix \(a, b = 1,1\). Denote the coefficient vector for expansion of $u$ in the $\bigWii$ OP basis up to degree $N$ by $\mathbf{u}$, and the coefficient vector for expansion of $f$ in the $\bigPii$ OP basis up to degree $N$ by $\mathbf{f}$. Since $f$ is known, we can obtain $\mathbf{f}$ using the quadrature rule above. In matrix-vector notation, our system hence becomes:
\begin{align}
    \laplacewii \mathbf{u} = \mathbf{f}
\end{align}
which can be solved to find $\mathbf{u}$.
In Figure \ref{fig:solution-poisson} we see the solution to the Poisson equation with zero boundary conditions given in (\ref{eqn:poisson}) in the half-disk $\Omega$. In Figure \ref{fig:solutionblocknorms} we show the norms of each block of calculated coefficients of the approximation for four right-hand sides of the Poisson equation with N = 200. The rate of decay in the coefficients is a proxy for the rate of convergence of the computed solution. \bstodo{Describe what plot tells us}

@@

Find \(u(x,y)\) given functions $v$, $f : \Omega \to \R$ such that:
\begin{align}
	\begin{cases}
    		\Delta u(x,y) + k^2 v(x,y) \; u(x,y) = f(x,y) \quad \text{in } \Omega \\
		u(x,y) = 0 \quad \text{on } \partial \Omega
	\end{cases}.
	\label{eqn:helmholtz}
\end{align}
where $k \in \R$, noting the imposition of zero Dirichlet boundary conditions on $u$.

We can tackle the problem as follows. Fix \(a, b = 1,1\). Denote the coefficient vector for expansion of $u$ in the $\bigWii$ OP basis up to degree $N$ by $\mathbf{u}$, and the coefficient vector for expansion of $f$ in the $\bigPii$ OP basis up to degree $N$ by $\mathbf{f}$. Since $f$ is known, we can obtain $\mathbf{f}$ using the quadrature rule above. We can obtain the matrix operator for the variable-coefficient function $v(x,y)$ by using the Clenshaw algorithm with matrix inputs as the Jacobi matrices ${J_x^{(0,0)}}^T, {J_y^{(0,0)}}^T$, yielding an operator matrix of the same dimension as the input Jacobi matries. We can denote the resulting operator acting on coefficients in the $\bigPoo$ space by $V({J_x^{(0,0)}}^T, {J_y^{(0,0)}}^T)$. In matrix-vector notation, our system hence becomes:
\begin{align}
    (\laplacewii + k^2 T^{(0,0)\to(1,1)} \: V({J_x^{(0,0)}}^T, {J_y^{(0,0)}}^T) \: T_W^{(1,1)\to(0,0)}) \mathbf{u} = \mathbf{f}
\end{align}
which can be solved to find $\mathbf{u}$. We can see the sparsity and structure of this matrix system in FIGURE with $v(x,y) = xy^2$ as an example. In FIGURE we see the solution to the inhomogenous variable-coefficient Helmholtz equation with zero boundary conditions given in (\ref{eqn:helmholtz}) in the half-disk $\Omega$, with  $v(x,y) = xy^2$ and $f(x,y) = xy e^x$.

@@

@@

Find \(u(x,y)\) given a function \(f(x,y)\) such that:
\begin{align}
	\begin{cases}
    		\Delta^2 u(x,y) = 0 \quad \text{in } \Omega \\
		u(x,y) = 0, \quad \frac{\partial u}{\partial n}(x,y) = 0 \quad \text{on } \partial \Omega
	\end{cases}.
	\label{eqn:biharmonic}
\end{align}
where $\Delta^2$ is the Biharmonic operator, noting the imposition of zero Dirichlet and Neumann boundary conditions on $u$. We can construct a matrix operator for $\Delta^2$ that will take coefficients in the space
$$
\bigW^{(2,2)}
$$
to coefficients in the space
$$
\bigP^{(2,2)}.
$$
Note that any function expanded in the $\bigW^{(2,2)}$ basis will satisfy both zero Dirichlet and zero Neumann boundary conditions on $\Omega$. The matrix operator for the Biharmonic operator is then given by
\begin{align}
	\biharmonic = \laplaceoo \: \laplacewtt.
\end{align}
In FIGURE we see the solution to the Biharmonic equation (\ref{eqn:biharmonic}) in the half-disk $\Omega$.
